---
title: "Historical outbreak frequency severity trend analysis"
author: "Amanda J. Meadows, PhD*, Nicole Stephenson, PhD, Nita K. Madhav, MSPH, and Ben Oppenheim, PhD"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(ggplot2)
library(nlme)
library(broom)
library(tidyverse)
library(data.table)
library(pander)
library(sf)
library(viridis)
theme_set(theme_bw())

```


# Selected Pathogens

Outbreak data was collected and analyzed for the following viral zoonotic pathogens:

```{r, results = 'asis'}
pathgrp_tab <- data.table("Path group" = c("Corona", "Filo", "Other"),
                          "Pathogens" = c("MERS Coronavirus, SARS-CoV", 
                                          "Marburg virus, Tai Forest ebolavirus, Sudan ebolavirus,
                                           Zaire ebolavirus, Bundibugyo ebolavirus",
                                          "Nipah virus, Machupo virus"))
cat(pander(pathgrp_tab, split.table = Inf))
```

## Data summary

Th number of outbreaks and deaths by pathogen group from 1960 through 2019:

```{r, results = 'asis'}
event_table <- fread("./data/spillover_data.csv")
lossDatYear <- event_table[, list(reported_deaths = sum(reported_deaths),
                                reported_cases = sum(reported_cases),
                                reported_events = uniqueN(Event_name)),
                         by = .(Event_start_year)]
lossDatYear[, Event_start_demidecade := Event_start_year - Event_start_year %% 5]

lossDatDemiDecade <- lossDatYear[,list(reported_deaths = sum(reported_deaths),
                                       reported_cases = sum(reported_cases),
                                       reported_events = sum(reported_events)),
                                 by = .(Event_start_demidecade)]
lossDatDemiDecade[, model_year := Event_start_demidecade - 1960]  
lossDat <- lossDatDemiDecade[Event_start_demidecade < 2020, ]
lossDatCovid <- lossDatDemiDecade
event_table[Pathogen %in% c("Machupo virus", "Nipah virus"), Pathogen_group := "Other"]
event_table[Pathogen %in% c("Sudan ebolavirus", "Zaire ebolavirus",
                            "Tai Forest ebolavirus", "Bundibugyo ebolavirus",
                            "Marburg virus"), Pathogen_group := "Filoviruses"]
event_table[Pathogen %in% c("MERS Coronavirus", "MERS-CoV",
                            "SARS-CoV"), Pathogen_group := "Epidemic coronaviruses"]
event_table[Pathogen %in% c("SARS Coronavirus 2"), Pathogen_group := "Excluded"]

event_tab <- event_table[, list(Outbreaks = .N, Deaths = sum(reported_deaths)), by = .(Pathogen_group)]
cat(pander(event_tab[Pathogen_group != "Excluded"]))
```

# Trend by demi decade {.tabset}

* Analyzing trend in events and deaths every 5 years from 1960 through 2019 (training data excludes Covid).
* Comparing a log fit and an exponential fit.

## Events{.tabset .tabset-pills}

* Trend in number of outbreaks per 5 year period (demidecade).

### Plot

```{r}
ggplot(lossDat, aes(x = Event_start_demidecade, y = reported_events)) + 
  geom_point() +
  stat_function(fun = function(x) 1.34756*exp(0.04234*(x-1950)), aes(color = "Exponential fit (AIC 73.87)"))+
  stat_function(fun = function(x) 10^(-50.515186 + 0.025738*(x)), aes(color = "Log fit (AIC 10.76)"))
```

### Exponential fit

```{r}
a_events <- 1
r_events <- 0.075

exp_events <- nls(reported_events ~ a*exp(r*model_year), 
                  data = lossDat,
                  start = list(a = a_events, r = r_events))
summary(exp_events)
```

#### AIC

```{r}
AIC(exp_events)
```

### Log fit

```{r}
logEvent <- lm(log10(reported_events)~ Event_start_demidecade, data = lossDat)
summary(logEvent)
```

#### AIC

```{r}
AIC(logEvent)
```

## Deaths {.tabset .tabset-pills}

### Plot

```{r}
ggplot(lossDat, aes(x = Event_start_demidecade, y = reported_deaths)) + 
  geom_point() +
  stat_function(fun = function(x) 62.44021*exp(0.07048*(x - 1950)), aes(color = "Exponential fit (AIC 229.85)"))+
  stat_function(fun = function(x) 10^(-90.55078 + 0.04657*(x)), aes(color = "Log fit (AIC 40.05)"))
```

### Exponential fit

```{r}
a_death <- 10
r_death <- 0.1

exp_deaths <- nls(reported_deaths ~ a*exp(r*model_year), 
                  data = lossDat[Event_start_demidecade > 1955],
                  start = list(a = a_death, r = r_death))
summary(exp_deaths)
```

#### AIC

```{r}
AIC(exp_deaths)
```

### Log fit

```{r}
# fit deaths and events vs time and make future predictions
logFit <- lm(log10(reported_deaths)~ Event_start_demidecade, data = lossDat)
summary(logFit)
```

#### AIC

```{r}
AIC(logFit)
```

### Log fit including Covid

```{r}
# fit deaths and events vs time and make future predictions
logFitCovid <- lm(log10(reported_deaths)~ Event_start_demidecade, data = lossDatCovid)
summary(logFitCovid)
```

## Conclusions

* Sticking with the log transformation model over the exponential fit because it is a better fit (supported by AIC).

# Future predictions {.tabset}

* 95% prediction interval

```{r}
# predicting out to 2030-2034
newDat <- data.table("Event_start_demidecade" = seq(1960, 2030, 5), "upr_demidecade" = seq(1964, 2034, 5))
newDat[, xlab := paste0(Event_start_demidecade, " - ", upr_demidecade)]

# add xlabels to data
lossDatDemiDecade <- merge(lossDatDemiDecade, newDat, by = "Event_start_demidecade")

# prediction and confidence intervals
## Deaths
predDat <- data.table(predict(logFit, newdata = newDat, interval = "predict", level = 0.95))
newDat <- cbind(newDat, setnames(predDat, c("log_fitDeath", "logPIlwr_Death", "logPIupr_Death")))
predDatCovid <- data.table(predict(logFitCovid, newdata = newDat, interval = "predict", level = 0.95))
newDat <- cbind(newDat, setnames(predDatCovid, c("log_fitDeathCovid", "logPIlwr_DeathCovid", "logPIupr_DeathCovid")))

# convert back to raw numbers
newDat[, fitDeath := 10^log_fitDeath]
newDat[, lwrDeathPI := 10^logPIlwr_Death]
newDat[, uprDeathPI := 10^logPIupr_Death]
newDat[, fitDeathCovid := 10^log_fitDeathCovid]
newDat[, lwrDeathPICovid := 10^logPIlwr_DeathCovid]
newDat[, uprDeathPICovid := 10^logPIupr_DeathCovid]

## events
predDatEvent <- data.table(predict(logEvent, newdata = newDat, interval = "predict", level = 0.97))
newDat <- cbind(newDat, setnames(predDatEvent, c("log_fitEvent", "logPIlwr_Event", "logPIupr_Event")))

newDat[, fitEvent := 10^log_fitEvent]
newDat[, lwrEventPI := 10^logPIlwr_Event]
newDat[, uprEventPI := 10^logPIupr_Event]
# outDat <- merge(lossDatDemiDecade, newDat, by = c("Event_start_demidecade", "upr_demidecade", "xlab"), all = T)
# fwrite(outDat, "data/figureData.csv")
```

## Deaths 

### Normal scale

```{r}
ggplot() + 
  geom_point(data= lossDatDemiDecade[Event_start_demidecade < 2020],
             aes(x = Event_start_demidecade, y = reported_deaths), size = 1) + 
  geom_point(data= lossDatDemiDecade[Event_start_demidecade > 2015],
             aes(x = Event_start_demidecade, y = reported_deaths), color = "red") +
  geom_line(data = newDat[Event_start_demidecade < 2020], 
            aes(x = Event_start_demidecade, y = fitDeath), color = "navy")+
  geom_line(data = newDat[Event_start_demidecade >= 2015], 
            aes(x = Event_start_demidecade, y = fitDeath), color = "navy", linetype = "dotted")+
  geom_line(data = newDat,
            aes(x = Event_start_demidecade, y = uprDeathPI), color = "steelblue",  linetype = "dashed")+
  geom_line(data = newDat, 
            aes(x = Event_start_demidecade, y = lwrDeathPI), color = "steelblue",  linetype = "dashed")+
  scale_x_continuous(breaks = seq(1960, 2030, by = 5), labels = newDat$xlab)+
  ylab("Reported deaths") + xlab("Year") + 
  coord_cartesian(ylim=c(0, 15000))+
  theme(axis.text.x = element_text(angle = 75, vjust = 0.5, hjust=0.5))
```

### Log10 scale

```{r}
comprss <- function(tx) {
      div <- findInterval(as.numeric(gsub("\\,", "", tx)),
         c(0, 1e3, 1e6, 1e9, 1e12) )  # modify this if negative numbers are possible
      paste(round( as.numeric(gsub("\\,","",tx))/10^(3*(div-1)), 2),
           c("","K","M","B","T")[div] )}
logybreak <- c(10^seq(0, 8, 1))
logylab <- comprss(logybreak)

fig1a_log <- ggplot() + 
  geom_point(data= lossDatDemiDecade[Event_start_demidecade < 2020],
             aes(x = Event_start_demidecade, y = reported_deaths)) + 
  # geom_point(data= lossDatDemiDecade[Event_start_demidecade >= 2020],
  #            aes(x = Event_start_demidecade, y = reported_deaths), color = "red") + 
 geom_line(data = newDat[Event_start_demidecade < 2020], 
            aes(x = Event_start_demidecade, y = fitDeath), color = "navy")+
  geom_line(data = newDat[Event_start_demidecade >= 2015], 
            aes(x = Event_start_demidecade, y = fitDeath), color = "navy", linetype = "dotted")+
  geom_line(data = newDat,
            aes(x = Event_start_demidecade, y = uprDeathPI), color = "steelblue",  linetype = "dashed")+
  geom_line(data = newDat, 
            aes(x = Event_start_demidecade, y = lwrDeathPI), color = "steelblue",  linetype = "dashed")+
  scale_x_continuous(breaks = seq(1960, 2030, by = 5), labels = newDat$xlab)+
  ylab("Deaths") + xlab("Year") + 
    #ylim(c(0, 1000000))+
  scale_y_log10(labels = logylab, breaks = logybreak) +
  theme(axis.text.x = element_text(angle = 75, vjust = 0.5, hjust=0.5))
fig1a_log
# ggsave(fig1a_log, filename = "./figures/log_reported_deaths.jpeg", dpi = 300, width = 5, height = 5, 
#        units = "in", device = "jpeg")
```

## Events 

### Normal scale

```{r}
ggplot() + 
  geom_point(data= lossDatDemiDecade[Event_start_demidecade < 2020],
             aes(x = Event_start_demidecade, y = reported_events)) + 
  geom_line(data = newDat[Event_start_demidecade < 2020], aes(x = Event_start_demidecade, y = fitEvent), color = "navy")+
  geom_line(data = newDat[Event_start_demidecade >= 2015], aes(x = Event_start_demidecade, y = fitEvent), 
            color = "navy", linetype = "dotted")+
  geom_line(data = newDat,
            aes(x = Event_start_demidecade, y = uprEventPI), color = "steelblue",  linetype = "dashed")+
  geom_line(data = newDat, 
            aes(x = Event_start_demidecade, y = lwrEventPI), color = "steelblue",  linetype = "dashed")+
  scale_x_continuous(breaks = seq(1960, 2030, by = 5), labels = newDat$xlab)+
  ylab("Reported events") + xlab("Year") + 
  theme(axis.text.x = element_text(angle = 75, vjust = 0.5, hjust=0.5)) + coord_cartesian(ylim=c(0, 100))
```

### Log10 scale

```{r}
ggplot() + 
  geom_point(data= lossDatDemiDecade[Event_start_demidecade < 2020],
             aes(x = Event_start_demidecade, y = reported_events)) + 
  geom_line(data = newDat[Event_start_demidecade < 2020], aes(x = Event_start_demidecade, y = fitEvent), color = "navy")+
   geom_line(data = newDat[Event_start_demidecade >= 2015], aes(x = Event_start_demidecade, y = fitEvent), color = "navy", linetype = "dotted")+
  geom_line(data = newDat,
            aes(x = Event_start_demidecade, y = uprEventPI), color = "steelblue",  linetype = "dashed")+
  geom_line(data = newDat, 
            aes(x = Event_start_demidecade, y = lwrEventPI), color = "steelblue",  linetype = "dashed")+
  scale_x_continuous(breaks = seq(1960, 2030, by = 5), labels = newDat$xlab)+
  ylab("Outbreaks") + xlab("Year") + 
  theme(axis.text.x = element_text(angle = 75, vjust = 0.5, hjust=0.5)) + scale_y_log10()
```


## Prediction table

The total number of events and deaths predicted to occur from 2020 to 2034 (±95% CI) according to the historical trend model.

```{r, results = 'asis'}
predTable <- newDat[Event_start_demidecade >= 2020, ][, list(Event_start_demidecade, 
                                                             Events_lwr = lwrEventPI, 
                                                             Events_upr = uprEventPI,
                                                             Events = fitEvent,
                                                             Deaths_lwr = lwrDeathPI,
                                                             Deaths_upr = uprDeathPI,
                                                             Deaths = fitDeath)]

predTable <- predTable[, list("Demidecade" = Event_start_demidecade, "Events" = paste0(round(Events,0), 
                                                " [",round(Events_lwr, 0)," - ", round(Events_upr, 0), "]"),
                              "Deaths" = paste0(format(round(Deaths, 0), big.mark = ","), 
                                                " [", round(Deaths_lwr, 0), " - ", format(round(Deaths_upr, 0), big.mark = ","), "]"))]
cat(pander(predTable))
```

## Prediction table (Covid-19 data)

The total number of events and deaths predicted to occur from 2020 to 2034 (±95% CI) according to the historical trend model if Covid-19 deaths up to 12/15/2021 are included in the training data.

```{r, results = 'asis'}
predTableCovid <- newDat[Event_start_demidecade >= 2020, ][, list(Event_start_demidecade, 
                                                             Deaths_lwr = lwrDeathPICovid,
                                                             Deaths_upr = uprDeathPICovid,
                                                             Deaths = fitDeathCovid)]

predTableCovid  <- predTableCovid[, list("Demidecade" = Event_start_demidecade, 
                                          "Deaths" = paste0(format(round(Deaths, 0), big.mark = ","), 
                                                " [", round(Deaths_lwr, 0), " - ", format(round(Deaths_upr, 0), big.mark = ","), "]"))]
cat(pander(predTableCovid))
```

# Contact

* Amanda Meadows: ameadows@metabiota.com